A Multi-Scale Analysis of 27,000 Urban Street Networks\footnote{This is a journal article manuscript currently under peer review.}

### 2017-03-26 latex/paper.tex
update intro and methods

\begin{abstract}
OpenStreetMap presents an emerging source of worldwide geospatial urban data useful to researchers. This paper uses the OSMnx software to automatically download and analyze 27,000 street networks from OpenStreetMap at metropolitan, municipal, and neighborhood scales---that is, every urbanized area, incorporated place, and Zillow neighborhood in the US. In turn it presents wide-ranging empirical findings on US urban form and street network characteristics. This paper addresses past limitations by conducting an analysis of street networks with an extremely large sample sizes, with clearly defined network definitions and extents for reproducibility, and using non-planar, directed graphs. Its analysis emphasizes urban street networks measures relevant to urban design and morphology, including structure, connectedness, centrality, and robustness. In total we cross-sectionally analyze the street networks of 497 urbanized areas, 19,655 cities and towns, and 6,857 neighborhoods. These sample sizes are larger than those in any previous study and illustrate the use of OSMnx and OpenStreetMap as a new urban form research platform.
\end{abstract}

\section{Introduction}
On May 20, 1862, Abraham Lincoln signed the Homestead Act into law, making land across the United States Midwest and Great Plains available for free to all applicants \cite{porterfield_homestead_2005}. Under its auspices over the next 70 years, the federal government distributed 270 million acres of public land (10\% of the entire US landmass) to private owners in the form of 1.6 million homesteads \cite{lee_kansas_1979, sherraden_inclusion_2005}. Towns with gridiron street networks sprang up rapidly across the Great Plains and Midwest, due to both the prevailing urban design paradigm of the day and the standardized rectilinear town plats used repeatedly to lay out instant new cities \cite{southworth_streets_1997}. Through path dependence, the spatial signatures of these land use laws, design paradigms, and planning instruments can still be seen today in these cities' urban forms and street networks.

A cross-sectional analysis of American urban form through its street networks at metropolitan, municipal, and neighborhood scales can reveal these artifacts and histories across the nation. Network analysis is a natural approach to the study of cities as growing, complex systems that self-organize to fill space with a capillary circulation network \cite{masucci_random_2009}. The empirical literature on street network analysis is growing ever richer, but suffers from some limitations---discussed in \cite{boeing_osmnx:_2017} and summarized here. First, sample sizes tend to be fairly small due to data availability, gathering, and processing constraints. Second, reproducibility is difficult when the dozens of decisions that go into analysis---such as spatial extents, topological simplification and correction, definitions of nodes and edges, etc.---are ad hoc or only partly reported. Third, and related to the first two, studies frequently oversimplify to planar or undirected primal graphs for tractability, or use dual graphs despite the loss of geographic and metric information.

This paper addresses these limitations by conducting an analysis of street networks at multiple scales, with large sample sizes, with clearly defined network definitions and extents for reproducibility, and using non-planar, directed graphs. In particular, it examines urban street networks---represented here as primal, non-planar, weighted multidigraphs with possible self-loops---focusing on structure, connectedness, centrality, and resilience. Most studies in the street network literature that conduct topological and/or metric analysis tend to have sample sizes ranging around 5 to 50 networks. This paper instead conducts a large analysis of over 27,000 urban street networks at multiple overlapping scales across the United States. Namely, it examines the street networks of every US incorporated city and town, urbanized area, and Zillow-defined neighborhood. In total, this study presented uses OSMnx \cite{boeing_osmnx:_2017} to download, construct, and analyze 497 urbanized areas' street networks, 19,655 cities' and towns' street networks, and 6,857 neighborhoods' street networks. It uses these street networks to conduct four analyses: at the metropolitan scale, at the municipal scale, at the neighborhood scale, and with a case study looking deeper at the neighborhood-scale street networks in the city of San Francisco.

This paper is organized as follows. In the next section, it presents the data sources, tools, and methods used to collect, construct, and analyze these street networks. Then it presents findings of the analyses at three spatial scales: metropolitan, municipal, and neighborhood. Finally, it concludes with a discussion of these findings and their implications for street network analysis, urban form complexity, and city planning.

\section{Methodology}
This study uses OSMnx to download, construct, correct, analyze, and visualize street networks at metropolitan, municipal, and neighborhood scales \cite{boeing_osmnx:_2017}. To define the study sites and their spatial boundaries, this study uses three sets of geometries. The first defines the metropolitan-scale study sites using the 2016 national TIGER/Line shapefile of US census bureau urban areas, released 19 August 2016. Census-defined urban areas comprise a set of census tracts that meet a minimum density threshold \cite{u.s._census_bureau_2010_2010}. In particular, we retain only the urbanized areas subset in this data set (i.e., areas with greater than 50,000 population), discarding the smaller urban clusters subset. The second set of geometries defines the municipal-scale study sites using 51 TIGER/Line shapefiles (again, 2016) of US census bureau places for all 50 states plus the District of Columbia. In particular, we discard the subset of census-designated places in this data set, instead retaining all the incorporated cities and towns in the United States.

@@

For all of these geometries, we use OSMnx to download the (drivable, public) street network contained within the geometry. First it buffers each geometry by 0.5 km, then downloads the street network within this buffered geometry. Next it constructs a street network from this data, corrects the topology, calculates street counts per node (this ensures that intersections are not considered cul-de-sacs simply because an incident edge connects to a node outside the desired polygon), then truncates the network to the original, desired polygon. OSMnx saves each of these networks to disk as GraphML and shapefiles. Finally, it automatically calculates metric and topological measures for each. However, in a planar graph, topological and metric properties are somewhat interrelated \cite{masucci_random_2009}.

@@

In total, this study cross-sectionally analyzes 27,009 networks: 497 urbanized areas' street networks, 19,655 cities' and towns' street networks, and 6,857 neighborhoods' street networks. These sample sizes are larger than those in any previous study. In the following sections, we present the findings of these analyses at the metropolitan scale, the municipal scale, the neighborhood scale, and through a case study looking deeper at the neighborhood-scale street networks in the city of San Francisco.

@@

\begin{figure}
\includegraphics[width=0.5\textwidth]{media/fig02.png}
\caption{Distribution of node types in 9 urbanized areas, with number of streets emanating from the node on the $x$-axis and proportion of nodes of this type on the $y$-axis }
\label{fig02}
\end{figure}

@@

The distribution of node types (i.e., intersections and dead ends) provides a clear indicator of network connectedness. The typical urbanized area has 2.8 streets per intersection on average: lots of 3-way intersections, fewer cul-de-sacs, and even fewer 4-way intersections. The grid-like San Angelo, TX urbanized area has the most streets per node (3.2) on average, and (outside of Puerto Rico, which contains the seven lowest urbanized areas) the sprawling, disconnected Lexington Park, MD urbanized area has the fewest (2.2). These two urban areas fit the trend seen in the spatial distribution across the US in Figure \ref{fig01}: urbanized areas in the great plains and Midwest have particularly high numbers of streets per node on average, indicating more grid-like, connected networks. Cities in the southern and western US tend to have fewer streets per node, reflecting more dead-ends and a disconnected network. This finding is discussed in more detail in the upcoming section.

In the typical urbanized area, 18\% of nodes are 4-way intersections, 59\% are 3-way intersections, and 21\% are dead-ends. However, this distribution varies somewhat between urbanized areas. Examining a small sample of 9 urbanized areas, chosen to maximize variance, reveals this in clearer detail. In Figure \ref{fig02}, urban Atlanta and Chattanooga have very high proportions of dead-ends, each over 30\% of all nodes, and very few 4-way intersections, indicating a disconnected street pattern. The urban areas of Phoenix, Boston, Detroit, and Chattanooga have particularly high proportions of 3-way intersections, each over 60\%, indicating a prevalence of T-intersections. Conversely, Chicago, New Orleans, Duluth, and Lubbock have high proportions of 4-way intersections, indicating more grid-like connected networks. But what is perhaps most notable about Figure \ref{fig02} is that these nine urbanized areas, despite being chosen to maximize variance, are overwhelmingly similar to each other. At the metropolitan scale, every large American urban agglomeration is characterized by a preponderance of 3-way intersections.

@@

Due to the substantial variation in urbanized area size, from 25 to 9,000 km\textsuperscript{2}, the preceding analysis covers a wide swath of metropolitan place types. To better compare apples-to-apples, Table \ref{table02} focuses on the 30 largest urban areas cross-sectionally to examine how their metric and topological measures compare. This provides more consistent spatial scales and extents, while offering a window into the similarities and differences in the built forms of America's largest urban agglomerations.

Among these urban areas, Milwaukee has the least circuitous network (6\% more circuitous than straight-line edges would be), and Orlando has the most (12\%). San Juan and Atlanta have the fewest streets per node on average (2.36 and 2.45, respectively), while Milwaukee has the most (3.03). Cincinnati has both the lowest intersection density (18/km\textsuperscript{2}) and street density (6.1 km/km\textsuperscript{2}) while Denver has the highest intersection density (40.6/km\textsuperscript{2}) and Miami and Los Angeles have the highest street density (10.6 km/km\textsuperscript{2}, apiece). In other words, Cincinnati has a particularly coarse-grained network with few connections and paths. This can also be seen in the average street segment length, a proxy for block size: Cincinnati has the second highest (186 m), bested only by Cleveland (198 m). In contrast, the two lowest are Denver's 138-meter average and San Juan's 131-meter average.

>>>>

\begin{abstract}

OpenStreetMap is an emerging source of worldwide geospatial data useful to researchers. This paper uses the OSMnx software to automatically download and analyze 27,000 street networks from OpenStreetMap at metropolitan, municipal, and neighborhood scales---that is, every urbanized area, incorporated place, and Zillow neighborhood in the US. In turn it presents wide-ranging empirical findings on US urban form and street network characteristics. This paper addresses past limitations by conducting an analysis of street networks with an extremely large sample sizes, with clearly defined network definitions and extents for reproducibility, and using non-planar, directed graphs. Its analysis emphasizes urban street network measures relevant to urban design and morphology, including structure, connectedness, centrality, and robustness. In total we cross-sectionally analyze the street networks of 497 urbanized areas, 19,655 cities and towns, and 6,857 neighborhoods. These sample sizes are larger than those in any previous study and illustrate the use of OSMnx and OpenStreetMap as a new urban form research platform.

\end{abstract}

\section{Introduction}

On May 20, 1862, Abraham Lincoln signed the Homestead Act into law, making land across the United States Midwest and Great Plains available for free to all applicants \cite{porterfield_homestead_2005}. Under its auspices over the next 70 years, the federal government distributed 270 million acres of public land (10\% of the entire US landmass) to private owners in the form of 1.6 million homesteads \cite{lee_kansas_1979, sherraden_inclusion_2005}. Towns with gridiron street networks sprang up rapidly across the Great Plains and Midwest, due to both the prevailing urban design paradigm of the day and the standardized rectilinear town plats used repeatedly to lay out instant new cities \cite{southworth_streets_1997}. Through path dependence, the spatial signatures of these land use laws, design paradigms, and planning instruments can still be seen today in these cities' urban forms and street networks.

A cross-sectional analysis of American urban form can reveal these artifacts and histories through street networks at metropolitan, municipal, and neighborhood scales. Network analysis is a natural approach to the study of cities as complex systems \cite{masucci_random_2009}. The empirical literature on street network analysis is growing ever richer, but suffers from some limitations---discussed in detail in \cite{boeing_osmnx:_2017} and summarized here. First, sample sizes have tended to be fairly small due to data availability, gathering, and processing constraints. Second, reproducibility has been difficult when the dozens of decisions that go into analysis---such as spatial extents, topological simplification and correction, definitions of nodes and edges, etc.---are ad hoc or only partly reported. Third, and related to the first two, studies frequently oversimplify to planar or undirected primal graphs for tractability, or use dual graphs despite the loss of geographic and metric information.

This paper addresses these limitations by conducting an urban morphological analysis of street networks at multiple scales, with large sample sizes, with clearly defined network definitions and extents for reproducibility, and using non-planar, directed graphs. In particular, it examines urban street network---represented here as primal, non-planar, weighted multidigraphs with possible self-loops---focusing on structure, connectedness, centrality, and resilience. Most studies in the street network literature that conduct topological and/or metric analysis tend to have sample sizes ranging around 10 to 50 networks. This study instead analyzes over 27,000 urban street networks at multiple overlapping scales across the United States. Namely, it examines the street networks of every incorporated city and town (n=19,655), census urbanized area (n=497), and Zillow-defined neighborhood (n=6,857) in the US. To do so, it uses OSMnx \cite{boeing_osmnx:_2017}---a new street network research toolkit---to download, construct, and analyze these street networks at metropolitan, municipal, and neighborhood scales.

The purpose of this paper is threefold. First, it demonstrates a new methodology for easily and consistently acquiring, constructing, and analyzing large samples of street networks as nonplanar directed graphs. Second, it presents empirical findings of descriptive urban morphology for the street networks of every city, urbanized area, and Zillow neighborhood in the US. Third, it investigates with large sample sizes some claims previously made in the research literature with small sample sizes. This paper is organized as follows. In the next section, it discusses the data sources, tools, and methods used to collect, construct, and analyze these street networks. Next, it presents findings of the analyses at metropolitan, municipal, and neighborhood spatial scales. Finally, it concludes with a discussion of these findings and their implications for street network analysis, urban morphology, and city planning.

\section{Methodology}

A network (also called a \emph{graph} in mathematics) comprises a set of nodes connected to one another by a set of edges. Street networks can be conceptualized as primal, directed, nonplanar graphs. A primal street network represents intersections as nodes and street segments as edges. A directed network has directed edges: that is, edge $uv$ points one-way from node $u$ to node $v$, but there need not exist a reciprocal edge $vu$. A planar network can be represented in two dimensions with its edges intersecting only at nodes. Most street networks are nonplanar---due to grade-separated expressways, overpasses, bridges, and tunnels---but most quantitative studies of urban street networks represent them as planar (e.g., Buhl et al. 2006; Cardillo et al. 2006; Barthélemy and Flammini 2008; Masucci et al. 2009; Strano et al. 2013) for tractability because bridges and tunnels are reasonably uncommon in some cities. Planarity may reasonably model the street networks of old European town centers, but poorly models the street network in more modern cities like Los Angeles with their grade-separated expressways, bridges, and tunnels. Thus, a planar simplification could yield inaccurate analyses.

This study uses OSMnx to download, construct, correct, analyze, and visualize street networks at metropolitan, municipal, and neighborhood scales \cite{boeing_osmnx:_2017}. OSMnx is a Python-based research tool that easily downloads OpenStreetMap data for any place in the world and constructs it into a spatially-embedded graph-theoretic object for analysis and visualization. OpenStreetMap is a collaborative worldwide mapping project that makes its spatial data available via various APIs. These data are of high quality and compare well to CIA World Factbook estimates and US Census TIGER/Line data (Haklay 2010; Over et al. 2010; Zielstra and Hochmair 2011; Maron 2015). In 2007, OpenStreetMap imported the 2005 TIGER/Line roads data, and since then many community-led corrections and improvements have been made (Willis 2008). Many of these additions go beyond TIGER/Line's scope, including passageways between buildings, footpaths through parks, bike paths, and detailed feature attributes, such as finer-grained street classifiers, speed limits, etc.

To define the study sites and their spatial boundaries, this study uses three sets of geometries. The first defines the metropolitan-scale study sites using the 2016 national TIGER/Line shapefile of US census bureau urban areas, released 19 August 2016. Census-defined urban areas comprise a set of census tracts that meet a minimum density threshold \cite{u.s._census_bureau_2010_2010}. In particular, we retain only the urbanized areas subset in this data set (i.e., areas with greater than 50,000 population), discarding the smaller urban clusters subset. The second set of geometries defines the municipal-scale study sites using 51 TIGER/Line shapefiles (again, 2016) of US census bureau places for all 50 states plus the District of Columbia. In particular, we discard the subset of census-designated places in this data set, instead retaining all the incorporated cities and towns in the United States.

@@

For all of these geometries, we use OSMnx to download the (drivable, public) street network contained within the geometry. First it buffers each geometry by 0.5 km, then downloads the street network within this buffered geometry. Next it constructs a street network from this data, corrects the topology, calculates street counts per node (this ensures that intersections are not considered cul-de-sacs simply because an incident edge connects to a node outside the desired polygon), then truncates the network to the original, desired polygon. OSMnx saves each of these networks to disk as GraphML and shapefiles. Finally, it automatically calculates metric and topological measures for each. Such measures extend the toolkit commonly used by urban morphologists (Talen 2003). Brief descriptions of these measures are discussed here, but extended definitions and algorithms can be found in Newman (2010) and Barthélemy (2011).

@@

In total, this study cross-sectionally analyzes 27,009 networks: 497 urbanized areas' street networks, 19,655 cities' and towns' street networks, and 6,857 neighborhoods' street networks. These sample sizes are larger than those in any previous similar study. In the following sections, we present the findings of these analyses at the metropolitan scale, the municipal scale, the neighborhood scale, and through a case study looking deeper at the neighborhood-scale street networks in the city of San Francisco.

@@

\begin{figure}
\includegraphics[width=0.75\textwidth]{media/fig02.png}
\caption{Distribution of node types in 9 urbanized areas, with number of streets emanating from the node on the $x$-axis and proportion of nodes of this type on the $y$-axis }
\label{fig02}
\end{figure}

@@

The distribution of node types (i.e., intersections and dead ends) provides a clear indicator of network connectedness. The typical urbanized area has 2.8 streets per intersection on average: lots of 3-way intersections, fewer cul-de-sacs, and even fewer 4-way intersections. The grid-like San Angelo, TX urbanized area has the most streets per node (3.2) on average, and (outside of Puerto Rico, which contains the seven lowest urbanized areas) the sprawling, disconnected Lexington Park, MD urbanized area has the fewest (2.2). These two urbanized areas fit the trend seen in the spatial distribution across the US in Figure \ref{fig01}: urbanized areas in the great plains and Midwest have particularly high numbers of streets per node on average, indicating more grid-like, connected networks. Cities in the southern and western US tend to have fewer streets per node, reflecting more dead-ends and a disconnected network. This finding is discussed in more detail in the upcoming section.

In the typical urbanized area, 18\% of nodes are 4-way intersections, 59\% are 3-way intersections, and 21\% are dead-ends. However, this distribution varies somewhat between urbanized areas. Examining a small sample of 9 urbanized areas, chosen to maximize variance, reveals this in clearer detail. In Figure \ref{fig02}, urban Atlanta and Chattanooga have very high proportions of dead-ends, each over 30\% of all nodes, and very few 4-way intersections, indicating a disconnected street pattern. The urbanized areas of Phoenix, Boston, Detroit, and Chattanooga have particularly high proportions of 3-way intersections, each over 60\%, indicating a prevalence of T-intersections. Conversely, Chicago, New Orleans, Duluth, and Lubbock have high proportions of 4-way intersections, indicating more grid-like connected networks. But what is perhaps most notable about Figure \ref{fig02} is that these nine urbanized areas, despite being chosen to maximize variance, are overwhelmingly similar to each other. At the metropolitan scale, every large American urban agglomeration is characterized by a preponderance of 3-way intersections.

@@

Due to the substantial variation in urbanized area size, from 25 to 9,000 km\textsuperscript{2}, the preceding analysis covers a wide swath of metropolitan place types. To better compare apples-to-apples, Table \ref{table02} focuses on the 30 largest urbanized areas cross-sectionally to examine how their metric and topological measures compare. This provides more consistent spatial scales and extents, while offering a window into the similarities and differences in the built forms of America's largest urban agglomerations.

Among these urbanized areas, Milwaukee has the least circuitous network (6\% more circuitous than straight-line edges would be), and Orlando has the most (12\%). San Juan and Atlanta have the fewest streets per node on average (2.36 and 2.45, respectively), while Milwaukee has the most (3.03). Cincinnati has both the lowest intersection density (18/km\textsuperscript{2}) and street density (6.1 km/km\textsuperscript{2}) while Denver has the highest intersection density (40.6/km\textsuperscript{2}) and Miami and Los Angeles have the highest street density (10.6 km/km\textsuperscript{2}, apiece). In other words, Cincinnati has a particularly coarse-grained network with few connections and paths. This can also be seen in the average street segment length, a proxy for block size: Cincinnati has the second highest (186 m), bested only by Cleveland (198 m). In contrast, the two lowest are Denver's 138-meter average and San Juan's 131-meter average.
